>

### 2019-04-22 paper/spinifex-paper.tex
introduction revised

% !TeX root = RJwrapper.tex
\title{spinifex: A Package for Creating an Animation of a Manually Controlled
Tour of Linear Projections of Multivariate Data, to Explore the
Sensitivity of Structure to Specific Variables}
\author{by Nicholas Spyrison, Dianne Cook}

@@

\abstract{%
Dynamic linear projections of multivariate data collectively known as
\textbf{tours} provide an important tool for exploring mulitvariate data
and models. The R package \textbf{tourr} provides functions for several
types of tours: grand, guided, little, local and frozen. Each of these
can be viewed dynamically, or saved into a data object for animation.
This paper describes a new package, \textbf{spinifex}, to provide a
manual tour, that allows the coefficient for a single variable to be
varied, with the result displayed using animation, with the packages,
\textbf{plotly} and \textbf{gganimate}. By varying the coefficient of a
single variable, it is possible to explore the sensitivity of structure
in the projection to that variable. It is particularly useful when used
with a projection pursuit guided tour to simplify and understand the
solution. The manual tour is applied to a problem from particle physics.
}

@@

Exploring multivariate spaces is a challenging task and increasingly so
as dimensionality increases. Traditionally, static projections display
an embedding of multivariate data in two dimensions. Such static views
cannot provide depth cues or convey the information of other
orientations. Alternatively, tours are a class of dynamic linear
projections viewed as an animation while the orientation of the
projection is rotated. Tours are a crucial tool extending the
dimensionality of visualization which is increasingly important as
datasets and models exist in increasing cardinality. The package
\CRANpkg{tourr} (Wickham et al.
\protect\hyperlink{ref-wickham_tourr_2011}{2011}) builds a platform for
generating tour paths and applying to a wide number of display
geometrics in \textbf{base} graphics.

While there are numerous methods that generate tour paths, this research
focuses on the manual tour. The manual tour was described in Cook and
Buja (\protect\hyperlink{ref-cook_manual_1997}{1997}) and allows a user
to control the projection coefficients of a selected variable in a 2D
projection. The manipulation of these coefficients allows the analyst to
explore their sensitivity to the structure within the projection. As
manual tours operate on only one variable at a time, they particularly
useful once a feature of interest has been identified. One way to
identify ``interesting'' features is with the use of a guided tour
(Hurley and Buja \protect\hyperlink{ref-hurley_analyzing_1990}{1990}).
Guided tours select a very specific path, that which approaches a
projection that optimizes an objective-function, in a manner similar to
simulated annealing (Kirkpatrick, Gelatt, and Vecchi
\protect\hyperlink{ref-kirkpatrick_optimization_1983}{1983}). The direct
optimization of a function allows guided tours to rapidly identify
interesting projection features given the relatively large
parameter-space. After a projection of interest is identified an analyst
can then use the ``finer brush'' of the manual tour, by controlling the
contributions of individual variables to explore the sensitivity they
have on the structure within the projection.

This paper offers an implementation of manual tours with the package
\CRANpkg{spinifex}. This package also extends the graphics display of
\textbf{tourr} to the animation frameworks offered by \CRANpkg{plotly}
(Sievert \protect\hyperlink{ref-sievert_plotly_2018}{2018}) and
\CRANpkg{gganimate} (Pedersen and Robinson
\protect\hyperlink{ref-pedersen_gganimate:_2019}{2019}). The algorithm
used to perform a radial manual tour is demonstrated on a toy data set.
Followed by an application and discussion of the radial manual tour on
an aggregation of high energy physics experiments in a large
multivariate space.

The paper is organized as follows. Section \ref{sec:algorithm} explains
the algorithm using a toy dataset. Section \ref{sec:display} discusses
the display of the animation after the path has been generated. Package
functionality and code usage following the order applied in the
algorithm follows in section \ref{sec:usage}. Section
\ref{sec:application} illustrates how this can be used for sensitivity
analysis applied to contemporary high energy physics. Then section
\ref{sec:summary} summarizes this paper. Lastly, section
\ref{sec:discussion} discusses possible future research.

@@

\hypertarget{sec:summary}{%
\subsection{Summary}\label{sec:summary}}

@@

\hypertarget{sec:discussion}{%
\section{Discussion}\label{sec:discussion}}

@@

@@

\leavevmode\hypertarget{ref-hurley_analyzing_1990}{}%
Hurley, C., and A. Buja. 1990. ``Analyzing High-Dimensional Data with
Motion Graphics.'' \emph{SIAM Journal on Scientific and Statistical
Computing} 11 (6): 1193--1211. \url{https://doi.org/10.1137/0911068}.

>>>>

% !TeX root = RJwrapper.tex
\title{spinifex: An R Package for Creating a Manual Tour of Low-dimensional
Projections of Multivariate Data}
\author{by Nicholas Spyrison, Dianne Cook}

@@

\abstract{%
Dynamic low-dimensional linear projections of multivariate data
collectively known as \textbf{tours} provide an important tool for
exploring mulitvariate data and models. The R package \textbf{tourr}
provides functions for several types of tours: grand, guided, little,
local and frozen. Each of these can be viewed dynamically, or saved into
a data object for animation. This paper describes a new package,
\textbf{spinifex}, to provide a manual tour, that allows the coefficient
for a single variable to be controlled. The variable is rotated fully
into the projection with a coefficient of 1 or -1, or completely out of
the projection with a coefficient of 0. The resulting sequence of
projections can be displayed using animation, with functions from either
the \textbf{plotly} and \textbf{gganimate} packages. By varying the
coefficient of a single variable, it is possible to explore the
sensitivity of structure in the projection to that variable. It is
particularly useful when used with a projection pursuit guided tour to
simplify and understand the solution. The use of the manual tour is
illustrated using a problem from particle physics.
}

@@

Exploring multivariate spaces is a challenging task, increasingly so as
dimensionality increases. Traditionally, static low-dimensional
projections are used to display multivariate data in two dimensions,
such as principal component analysis, linear discriminant spaces or
projection pursuit. These are useful for finding relationships between
multiple variables, but they are limited because they show only a
glimpse of the high-dimensional space. An alternative approach, is to
use a tour of dynamic linear projections, to look at many different
low-dimensional proections. Tours can be considered to extend the
dimensionality of visualization, which is important as data and models
exist in more than 3D. The package \CRANpkg{tourr} (Wickham et al.
\protect\hyperlink{ref-wickham_tourr_2011}{2011}) provides a platform
for generating tours. It has many types of tours available, and many
types of displays possible. A user can make a grand, guided, little,
local or frozen tour, and display the resulting projected data as a
scatterplot, density plot, histogram, or even as Chernoff faces if the
projection dimension is more than 3.

This work adds a manual tour to the collection. The manual tour was
described in Cook and Buja
(\protect\hyperlink{ref-cook_manual_1997}{1997}) and allows a user to
control the projection coefficients of a selected variable in a 2D
projection. The manipulation of these coefficients allows the analyst to
explore their sensitivity to the structure within the projection. As
manual tours operate on only one variable at a time, they are
particularly useful once a feature of interest has been identified.

One way to identify ``interesting'' features is with the use of a guided
tour (Cook et al. \protect\hyperlink{ref-cook_grand_1995}{1995}). Guided
tours select a very specific path, that which approaches a projection
that optimizes an objective-function. The optimization is conducted in a
manner similar to simulated annealing (Kirkpatrick, Gelatt, and Vecchi
\protect\hyperlink{ref-kirkpatrick_optimization_1983}{1983}). The direct
optimization of a function allows guided tours to rapidly identify
interesting projection features given the relatively large
parameter-space. After a projection of interest is identified an analyst
can then use the ``finer brush'' of the manual tour, by controlling the
contributions of individual variables to explore the sensitivity they
have on the structure visible in the projection.

The paper is organized as follows. Section \ref{sec:algorithm} describes
the algorithm used to perform a radial manual tour as implemented in the
package \CRANpkg{spinifex}. Section \ref{sec:display} explains how to
generate an animation of the manual tour using the animation frameworks
offered by \CRANpkg{plotly} (Sievert
\protect\hyperlink{ref-sievert_plotly_2018}{2018}) and
\CRANpkg{gganimate} (Pedersen and Robinson
\protect\hyperlink{ref-pedersen_gganimate:_2019}{2019}). Package
functionality and code usage following the order applied in the
algorithm follows in section \ref{sec:usage}. Section
\ref{sec:application} illustrates how this can be used for sensitivity
analysis applied to multivariate data collected on high energy physics
experiments (Wang et al.
\protect\hyperlink{ref-wang_mapping_2018}{2018}). Section
\ref{sec:discussion} summarizes this paper and discusses potential
future directions.

@@

\hypertarget{sec:discussion}{%
\subsection{Discussion}\label{sec:discussion}}

@@

@@

\leavevmode\hypertarget{ref-cook_grand_1995}{}%
Cook, Dianne, Andreas Buja, Javier Cabrera, and Catherine Hurley. 1995.
``Grand Tour and Projection Pursuit.'' \emph{Journal of Computational
and Graphical Statistics} 4 (3): 155.
\url{https://doi.org/10.2307/1390844}.

@@

### 2019-04-22 paper/spinifex_paper.Rmd
introduction revised

---
title: > 
  spinifex: A Package for Creating an Animation of a Manually Controlled Tour of Linear Projections of Multivariate Data, to Explore the Sensitivity of Structure to Specific Variables
author:
  - name: Nicholas Spyrison
    affiliation: Monash University 
    address:
    - Faculty of Information Technology
    email:  Nicholas.Spyrison@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
abstract: >
  Dynamic linear projections of multivariate data collectively known as **tours** provide an important tool for exploring mulitvariate data and models. The R package **tourr** provides functions for several types of tours: grand, guided, little, local and frozen. Each of these can be viewed dynamically, or saved into a data object for animation. This paper describes a new package, **spinifex**, to provide a manual tour, that allows the coefficient for a single variable to be varied, with the result displayed using animation, with the packages, **plotly** and **gganimate**. By varying the coefficient of a single variable, it is possible to explore the sensitivity of structure in the projection to that variable. It is particularly useful when used with a projection pursuit guided tour to simplify and understand the solution. The manual tour is applied to a problem from particle physics.

@@

<!-- Tours -->
Exploring multivariate spaces is a challenging task and increasingly so as dimensionality increases. Traditionally, static projections display an embedding of multivariate data in two dimensions. Such static views cannot provide depth cues or convey the information of other orientations. Alternatively, tours are a class of dynamic linear projections viewed as an animation while the orientation of the projection is rotated. Tours are a crucial tool extending the dimensionality of visualization which is increasingly important as datasets and models exist in increasing cardinality. The package \CRANpkg{tourr} [@wickham_tourr_2011] builds a platform for generating tour paths and applying to a wide number of display geometrics in **base** graphics.

<!-- Manual tour -->
While there are numerous methods that generate tour paths, this research focuses on the manual tour. The manual tour was described in @cook_manual_1997 and allows a user to control the projection coefficients of a selected variable in a 2D projection. The manipulation of these coefficients allows the analyst to explore their sensitivity to the structure within the projection. As manual tours operate on only one variable at a time, they particularly useful once a feature of interest has been identified. One way to identify "interesting" features is with the use of a guided tour [@hurley_analyzing_1990]. Guided tours select a very specific path, that which approaches a projection that optimizes an objective-function, in a manner similar to simulated annealing [@kirkpatrick_optimization_1983]. The direct optimization of a function allows guided tours to rapidly identify interesting projection features given the relatively large parameter-space. After a projection of interest is identified an analyst can then use the "finer brush" of the manual tour, by controlling the contributions of individual variables to explore the sensitivity they have on the structure within the projection.

<!-- Contributions -->
This paper offers an implementation of manual tours with the package
\CRANpkg{spinifex}. This package also extends the graphics display of **tourr** to the animation frameworks offered by \CRANpkg{plotly} [@sievert_plotly_2018] and \CRANpkg{gganimate} [@pedersen_gganimate:_2019]. The algorithm used to perform a radial manual tour is demonstrated on a toy data set. Followed by an application and discussion of the radial manual tour on an aggregation of high energy physics experiments in a large multivariate space.

<!-- Paper outline -->
The paper is organized as follows. Section \ref{sec:algorithm} explains the algorithm using a toy dataset. Section \ref{sec:display} discusses the display of the animation after the path has been generated. Package functionality and code usage following the order applied in the algorithm follows in section \ref{sec:usage}. Section \ref{sec:application} illustrates how this can be used for sensitivity analysis applied to contemporary high energy physics. Then section \ref{sec:summary} summarizes this paper. Lastly, section \ref{sec:discussion} discusses possible future research.

@@

## Summary {#sec:summary}

@@

# Discussion {#sec:discussion}

>>>>

---
title: > 
  spinifex: An R Package for Creating a Manual Tour of Low-dimensional  Projections of Multivariate Data
author:
  - name: Nicholas Spyrison
    affiliation: Monash University 
    address:
    - Faculty of Information Technology
    email:  Nicholas.Spyrison@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
abstract: >
  Dynamic low-dimensional linear projections of multivariate data collectively known as **tours** provide an important tool for exploring mulitvariate data and models. The R package **tourr** provides functions for several types of tours: grand, guided, little, local and frozen. Each of these can be viewed dynamically, or saved into a data object for animation. This paper describes a new package, **spinifex**, to provide a manual tour, that allows the coefficient for a single variable to be controlled. The variable is rotated fully into the projection with a coefficient of 1 or -1, or completely out of the projection with a coefficient of 0. The resulting sequence of projections can be displayed using animation, with functions from either the **plotly** and **gganimate** packages. By varying the coefficient of a single variable, it is possible to explore the sensitivity of structure in the projection to that variable. It is particularly useful when used with a projection pursuit guided tour to simplify and understand the solution. The use of the manual tour is illustrated using a problem from particle physics.

@@

<!-- Tours -->
Exploring multivariate spaces is a challenging task, increasingly so as dimensionality increases. Traditionally, static low-dimensional projections are used to display multivariate data in two dimensions, such as principal component analysis, linear discriminant spaces or projection pursuit. These are useful for finding relationships between multiple variables, but they are limited because they show only a glimpse of the high-dimensional space. An alternative approach, is to use a tour of dynamic linear projections, to look at many different low-dimensional proections. Tours can be considered to extend the dimensionality of visualization, which is important as data and models exist in more than 3D. The package \CRANpkg{tourr} [@wickham_tourr_2011] provides a platform for generating tours. It has many types of tours available, and many types of displays possible. A user can make a grand, guided, little, local or frozen tour, and display the resulting projected data as a scatterplot, density plot, histogram, or even as Chernoff faces if the projection dimension is more than 3.

<!-- Manual tour -->
This work adds a manual tour to the collection. The manual tour was described in @cook_manual_1997 and allows a user to control the projection coefficients of a selected variable in a 2D projection. The manipulation of these coefficients allows the analyst to explore their sensitivity to the structure within the projection. As manual tours operate on only one variable at a time, they are particularly useful once a feature of interest has been identified.

One way to identify "interesting" features is with the use of a guided tour [@cook_grand_1995]. Guided tours select a very specific path, that which approaches a projection that optimizes an objective-function. The optimization is conducted in a manner similar to simulated annealing [@kirkpatrick_optimization_1983]. The direct optimization of a function allows guided tours to rapidly identify interesting projection features given the relatively large parameter-space. After a projection of interest is identified an analyst can then use the "finer brush" of the manual tour, by controlling the contributions of individual variables to explore the sensitivity they have on the structure visible in the projection.

<!-- Contributions 
This paper describes the algorithm used to perform a radial manual tour as implemented in the package \CRANpkg{spinifex}. It also describes how to generate an animation of the manual tour using the animation frameworks offered by \CRANpkg{plotly} [@sievert_plotly_2018] and \CRANpkg{gganimate} [@pedersen_gganimate:_2019]. The method is applied to multivariate data collected on high energy physics experiments (REF).
-->

<!-- Paper outline -->
The paper is organized as follows. Section \ref{sec:algorithm} describes the algorithm used to perform a radial manual tour as implemented in the package \CRANpkg{spinifex}. Section \ref{sec:display} explains how to generate an animation of the manual tour using the animation frameworks offered by \CRANpkg{plotly} [@sievert_plotly_2018] and \CRANpkg{gganimate} [@pedersen_gganimate:_2019]. Package functionality and code usage following the order applied in the algorithm follows in section \ref{sec:usage}. Section \ref{sec:application} illustrates how this can be used for sensitivity analysis applied to multivariate data collected on high energy physics experiments [@wang_mapping_2018]. Section \ref{sec:discussion} summarizes this paper and discusses potential future directions.

@@

## Discussion {#sec:discussion}

@@
